---
title: "sSNAPPY"
output: html_document
date: "2023-09-19"
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 7,
  fig.width = 10
)
```

```{r packages}
library(tidyverse)
library(yaml)
library(scales)
library(pander)
library(glue)
library(edgeR)
library(AnnotationHub)
library(ensembldb)
library(magrittr)
library(cowplot)
library(ggfortify)
library(cqn)
library(ggrepel)
library(DT)
library(randomcoloR)
library(ggforce)
library(sSNAPPY)
library(rvest)
library(xml2)
library(BSDA)
library(corrplot)
```


```{r options}
panderOptions("table.split.table", Inf)
panderOptions("big.mark", ",")
theme_set(theme_bw())
```

```{r config}
config <- here::here("config/config.yml") %>%
  read_yaml()
suffix <- paste0(config$tag)
sp <- config$ref$species %>%
  str_replace("(^[a-z])[a-z]*_([a-z]+)", "\\1\\2") %>%
  str_to_title()
```

```{r}
if (!dir.exists(here::here("docs/assets"))) dir.create(here::here("docs/assets"))
```


# Setup

## Gene Annotations

```{r ah}
ah <- AnnotationHub() %>%
  subset(rdataclass == "EnsDb") %>%
  subset(str_detect(description, as.character(config$ref$release))) %>%
  subset(genome == config$ref$build)
stopifnot(length(ah) == 1)
```

```{r ensDb}
ensDb <- ah[[1]]
genesGR <- read_rds(here::here("output/genesGR.rds"))
```

Gene annotations were again loaded from Ensembl Release `r ensemblVersion(ensDb)`.
The [previously defined](qc_aligned.html#Annotation_Setup) `GenomicRanges` object containing GC content and Gene Length was also loaded,
containing information for `r comma(length(genesGR))` genes.

## Read in data

```{r}
diag_cols <- readRDS("~/GSE800098/output/diag_cols.rds")
```

The filtered and normalised `DGEList` was loaded in. 

```{r importData}
dge <- here::here("output/dge.rds") %>%
  read_rds()
logCPM <- dge$counts %>%
  edgeR::cpm(log = TRUE)
```


# Retrieve KEGG pathways

To apply the `sSNAPPY` algorithm, pathway topologies of KEGG pathways were firstly retrieved. 

```{r gsTopology}
gsTopology <- retrieve_topology(database = "kegg", species = "hsapien")
```

`r length(gsTopology)` KEGG pathways were retrieved but as described in the PDE study, irrelevant pathways are removed to keep only the 235 relevant pathways.

```{r}
gsTokeep <- readRDS("~/20131906_HickeyT_JC_NormalBreast/output/kg_gsTokeep.rds")
gsTopology <- gsTopology %>%
    .[names(.) %in% gsTokeep]
```

# sSNAPPY

To compute the single-sample logFCs required, rownames of the logCPM matrix need to be converted to `entrez id` first.

```{r entrezId}
rownames(logCPM) <- mapIds(ensDb, rownames(logCPM), "ENTREZID", keytype = "GENEID")
# Remove genes that couldn't be matched to entrez IDs
logCPM <- logCPM[!is.na(rownames(logCPM)),]
```

Weighted single sample logFCs were computed from the logCPM matrix.

```{r weightedFC}
metadata <- dge$samples %>%
    dplyr::rename(treatment = treat) %>%
    droplevels() %>%
    mutate(
        treatment = relevel(treatment, ref = "Vehicle")
    )
weightedFC <- weight_ss_fc(logCPM, metadata, sampleColumn = "name", 
                           treatColumn = "treatment", 
                           groupBy = "rep")
```


Raw single sample perturbation scores for each treated sample each KEGG pathway were calculated. 
```{r ssPertScore}
gene_pert <- sSNAPPY::raw_gene_pert(weightedFC$weighted_logFC, gsTopology)
ssPertScore <- pathway_pert(gene_pert, weightedFC = weightedFC$weighted_logFC)
```

Null distribution of perturbation scores were generated through sample label permutations. 

```{r permutedScore}
set.seed(123)
permutedScore <- generate_permuted_scores(
    logCPM,
    gsTopology = gsTopology,
    weight = weightedFC$weight
)
write_rds(permutedScore, here::here("output/MDA453_permutedscore.rds"))

permutedScore <- read_rds(here::here("output/MDA453_permutedscore.rds"))
```

Since the sample size is fairly short (ie. 12 samples), the maximum number of possible permutation pairs is 132. The permuted distributions don't look very impressive. 

```{r hist_pl, fig.width=10, fig.height=5}
set.seed(123)
pl <- permutedScore %>%
    keep(~all(.!=0)) %>%
    .[sample(seq_along(.), 6)] %>%
    lapply(
      function(x){
        ggplot(mapping = aes(x)) + 
          geom_histogram() +
          xlab("Perturbation Score")
      }
    ) 
plot_grid(plotlist = pl, nrow = 2)
```

### Sample-wise

Raw perturbation scores were converted to robust z-scores based on the median and MAD derived from the null distributions. 
A permuted p-value was computed for each single-sample perturbation scores and corrected for multiple testings. 

No pathway was defined to be significantly perturbed at individual sample levels. 

```{r normalisedScores}
normalisedScores <- normalise_by_permu(permutedScore, ssPertScore)
normalisedScores %>%
    dplyr::filter(adjPvalue < 0.05)
```

### Average treatment effect

Using the sample-level perturbation scores obtained above, we can testing for group-level perturbation by applying t-tests across all treated samples.
Using a FDR cut-off of 0.05, only 2 pathways were considered to be significantly perturbed

```{r}
z_matrix <- normalisedScores %>%
  dplyr::select(robustZ, gs_name, sample) %>%
  # dplyr::filter(sample != "MDA2-D") %>%
  pivot_wider(names_from = "sample", values_from = "robustZ") %>%
  column_to_rownames("gs_name") %>%
  as.matrix()
sapply(rownames(z_matrix), function(x){
  test <- t.test(z_matrix[x,]) 
  data.frame(
    gs_name = x, 
    pvalue = test$p.value, 
    t_statistic = test$statistic
  )
}, simplify = FALSE) %>%
  bind_rows() %>%
  mutate(FDR = p.adjust(pvalue, "fdr")) %>%
  dplyr::filter(
    FDR < 0.05
  )
```
